---
title: "Hands-on Exercise 1"
author: "Teong Sze Ching"
date: "January 13, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Lesson 1: A Layered Grammar of Graphics: ggplot2 methods

## 1.0 Getting Started
Before we get started, it is important for us to ensure that the required R packages have been installed. If yes, we will load the R packages. If they have yet to be installed, we will install the R packages and load them onto R environment.
```{r}
pacman::p_load(tidyverse)
```

# 2.0 The Data
Loading the data to the R environment by using read_csv() function of readr package
```{r}
exam_data <- read_csv("chap01/data/Exam_data.csv")
```
The data in the file includes (1)Year end examination grades of a cohort of primary 3 students from a local school. (2) There are a total of seven attributes. Four of them are categorical data type and the other three are in continuous data type. The categorical attributes are: ID, CLASS, GENDER and RACE. The continuous attributes are: MATHS, ENGLISH and SCIENCE.

# 3.0 Visualization - Math Grades 
## 3.1 By Dot Plot
```{r}
ggplot(data=exam_data, aes(x=MATHS)) + geom_dotplot(binwidth=2.5, dotsize=0.5) + scale_y_continuous(NULL,breaks=NULL)

```

## 3.2 By Histogram
```{r}
ggplot(data=exam_data, aes(x=MATHS, fill=GENDER)) + geom_histogram(bins=20,color="grey30")

```

## 3.3 By Density
```{r}
ggplot(data=exam_data, aes(x=MATHS, color=GENDER)) + geom_density()

```

## 3.4 By Box Plot
```{r}
ggplot(data=exam_data, aes(y=MATHS, x=GENDER)) + geom_boxplot() + stat_summary(geom="point", fun="mean",color="red",size=4)

```
```{r}
ggplot(data=exam_data, aes(y=MATHS, x=GENDER)) + geom_boxplot() + geom_point(position="jitter",size=0.5)

```

## 3.5 By Violin Plot
```{r}
ggplot(data=exam_data, aes(y=MATHS, x=GENDER)) + geom_violin()

```

## 3.6 By Trellis Plot
```{r}
ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins=20) + facet_wrap(~ CLASS)

```

```{r}
ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins=20) + facet_grid(~ CLASS)

```

# 4.0 Visualization - Race
## 4.1 Race by Vertical Bar Chart
```{r}
ggplot(data=exam_data, aes(x=RACE)) + geom_bar()

```

## 4.2 Race by Horizontal Bar Chart
```{r}
ggplot(data=exam_data, aes(x=RACE)) + geom_bar() + coord_flip() +theme_minimal()

```

# 5.0 Visualization - Math and English Grades 
```{r}
ggplot(data=exam_data, aes(x=MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method=lm,linewidth=0.5)

```

## 5.1 Fixing scale of both Math and English grades from 0 to 100
```{r}
ggplot(data=exam_data, aes(x=MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method=lm,size=0.5) + coord_cartesian(xlim=c(0,100),ylim=c(0,100))

```