---
title: "Hands-on Exercise 2"
author: "Teong Sze Ching"
date: "January 18, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Lesson 2: Beyond ggplot2 Fundamentals

## 1.0 Getting Started
In this chapter, several ggplot2 extensions will be introduced to create more elegant and effective statistical graphics.

In this exercise, beside tidyverse, four R packages will be used. They are:
- ggrepel: an R package provides geoms for ggplot2 to repel overlapping text labels.
- ggthemes: an R package provides some extra themes, geoms, and scales for ‘ggplot2’.
- hrbrthemes: an R package provides typography-centric themes and theme components for ggplot2.
- patchwork: an R package for preparing composite figure created using ggplot2.

Code chunk below will be used to check if these packages have been installed and also will load them onto your working R environment.

```{r}
pacman::p_load(ggrepel, patchwork, ggthemes, tidyverse)
```

hrbrtheme will not be loaded because it is not available. Will use ggplot2 to replace it. Link: https://cran.r-project.org/web/packages/hrbrthemes/index.html

# 2.0 The Data
Loading the data to the R environment by using read_csv() function of readr package
```{r}
exam_data <- read_csv("chap02/data/Exam_data.csv")
```
The data in the file includes (1)Year end examination grades of a cohort of primary 3 students from a local school. (2) There are a total of seven attributes. Four of them are categorical data type and the other three are in continuous data type. The categorical attributes are: ID, CLASS, GENDER and RACE. The continuous attributes are: MATHS, ENGLISH and SCIENCE.

# 3.0 Visualization
## 3.1 Without GGrepel
```{r}
ggplot(data=exam_data, aes(x=MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method=lm, size=0.5) + geom_label(aes(label=ID), hjust=.5, vjust=-.5) + coord_cartesian(xlim=c(0,100), ylim=c(0,100)) + ggtitle("English scores versus Maths scores for Primary 3")

```

## 3.2 With GGrepel
```{r}
ggplot(data=exam_data, aes(x=MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method=lm, size=0.5) + geom_label_repel(aes(label=ID), fontface="bold") + coord_cartesian(xlim=c(0,100), ylim=c(0,100)) + ggtitle("English scores versus Maths scores for Primary 3")

```

## 3.3 Using GGplot2 theme
ggplot2 comes with eight built-in themes, they are: theme_gray(), theme_bw(), theme_classic(), theme_dark(), theme_light(), theme_linedraw(), theme_minimal(), and theme_void().
```{r}
ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins=20, boundary=100, color="grey25", fill="grey90") + theme_gray() + ggtitle("Distribution of Maths score")

```

### 3.3.1 Using theme_economist()
```{r}
ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins=20, boundary=100, color="grey25", fill="grey90") + theme_gray() + ggtitle("Distribution of Maths score") + theme_economist()

```

### 3.3.2 Using theme_minimal() from ggplot2 (replaced it with theme_ipsum() because this is not available in the current R version)
```{r}
ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins=20, boundary=100, color="grey25", fill="grey90") + theme_gray() + ggtitle("Distribution of Maths score") + theme_minimal(base_size = 15, base_family="Arial") + theme(axis.title = element_text(size = 18), panel.grid.major.x = element_blank())

```

## 3.4 Using Composite graph
### 3.4.1 1st Statistical Graph - Distribution of Maths score
```{r}
p1 <- ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins=20,boundary=100, color="grey25", fill="grey90") + coord_cartesian(xlim=c(0,100)) + ggtitle("Distribution of Maths scores")

```
```{r, echo=FALSE}
p1
```

### 3.4.2 2nd Statistical Graph - Distribution of English score
```{r}
p2 <- ggplot(data=exam_data, aes(x=ENGLISH)) + geom_histogram(bins=20, boundary=100, color="grey25", fill="grey90") + coord_cartesian(xlim=c(0,100)) + ggtitle("Distribution of English scores")

```
```{r, echo=FALSE}
p2
```


### 3.4.3 Combine 1st and 2nd graph - Scatterplot for English vs Maths score

```{r}
p3 <- ggplot(data=exam_data, aes(x=MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method=lm, size=0.5) + coord_cartesian(xlim=c(0,100), ylim=c(0,100)) + ggtitle("English scores versus Maths scores for Primary 3")

```
```{r, echo=FALSE}
p3
```

#### 3.4.3.1 Creating scatterplot with insert
```{r}
insert_element <- patchwork::inset_element
p3 + insert_element(p2, left=0.02, bottom=0.7, right=0.5, top=1)
```

### 3.4.4 Combine 2 ggplot2 graphs 
```{r}
p1+p2
```

### 3.4.5 Combine 3 or more ggplot2 graphs 
We can plot more complex composite by using appropriate operators. For example, the composite figure below is plotted by using:

“/” operator to stack two ggplot2 graphs,
“|” operator to place the plots beside each other,
“()” operator the define the sequence of the plotting.

```{r, fig.width=12, fig.height=8}
p1/p2 | p3
```

### 3.4.6 Creating a composite figure by using patchwork and ggtheme
```{r, fig.width=12, fig.height=8}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
