---
title: "Hands-on Exercise 5"
author: "Teong Sze Ching"
date: "February 8, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Lesson 5-A: Creating Ternary Plot with R

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## 2.0 Data

Loading the data to the R environment by using read_csv() function of readr package

```{r}
pop_data <- read.csv("chap05/data/respopagsex2000to2018.csv")

```

Using the mutate() function of dplyr package to derive three new measures, namely: young, active, and old:

```{r}
#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## 3.0 Plotting Ternary Diagram with R

### 3.1 Plotting a static ternary diagram

```{r}
#Building the static ternary plot
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) + geom_point()
```

```{r}
#Building the static ternary plot
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +  geom_point() + labs(title="Population structure, 2015") +  theme_rgbw()
```

### 3.2 Plotting an interactive ternary diagram

```{r}
# reusable function for creating annotation object
label <- function(txt) {list(text = txt, x = 0.1, y = 1, ax = 0, ay = 0, xref = "paper", yref = "paper", align = "center", font =list(family = "serif", size = 15, color = "white"), bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2)}

# reusable function for axis formatting
axis <- function(txt) {list(title = txt, tickformat = ".0%", tickfont = list(size = 10))}

ternaryAxes <- list(aaxis = axis("Young"), baxis = axis("Active"),  caxis = axis("Old"))

# Initiating a plotly visualization 
plot_ly(agpop_mutated, a = ~YOUNG, b = ~ACTIVE, c = ~OLD, color = I("black"), type = "scatterternary") %>%  layout(annotations = label("Ternary Markers"), ternary = ternaryAxes)
```

# Lesson 5-B: Visual Correlation Analysis

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

## 2.0 Data

Loading the data to the R environment by using read_csv() function of readr package

```{r}
wine <- read.csv("chap05/data/wine_quality.csv")
```

## 3.0 Building Correlation Matrix: pairs() method

### 3.1 Building a basic correlation matrix

```{r}
pairs(wine[,1:11])
```

```{r}
pairs(wine[,2:12])
```

### 3.2 Drawing the lower corner

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

```{r}
pairs(wine[,2:12], lower.panel = NULL)
```

### 3.3 Including with correlation coefficients

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12], upper.panel = panel.cor)
```

### 3.4 Visualising Correlation Matrix: ggcormat()

#### 3.4.1 The Basic Plot

```{r}
ggstatsplot::ggcorrmat(data = wine, cor.vars = 1:11)
```

```{r}
ggstatsplot::ggcorrmat(data = wine, cor.vars = 1:11, ggcorrplot.args = list(outline.color = "black", hc.order = TRUE, tl.cex = 10), title = "Correlogram for wine dataset", subtitle = "Four pairs are no significant at p < 0.05")
```

```{r}
ggplot.component = list(theme(text=element_text(size=5), axis.text.x = element_text(size = 8), axis.text.y = element_text(size = 8)))
```

### 3.4 Building Multiple Plots

```{r}
grouped_ggcorrmat(data = wine, cor.vars = 1:11, grouping.var = type, type = "robust", p.adjust.method = "holm", plotgrid.args = list(ncol = 2), ggcorrplot.args = list(outline.color = "black", hc.order = TRUE, tl.cex = 10), annotation.args = list(tag_levels = "a", title = "Correlogram for wine dataset", subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity", caption = "Dataset: UCI Machine Learning Repository"))
```

### 3.5 Visualising Correlation Matrix using corrplot Package

#### 3.5.1 Getting started with corrplot

```{r}
wine.cor <- cor(wine[, 1:11])
```

```{r}
corrplot(wine.cor)
```

#### 3.5.2 Working with visual geometrics

```{r}
corrplot(wine.cor, method = "ellipse") 
```

```{r}
corrplot(wine.cor, method = "ellipse", type="lower")
```

```{r}
corrplot(wine.cor, method = "ellipse", type="lower", diag = FALSE, tl.col = "black")
```

#### 3.5.3 Working with mixed layout

```{r}
corrplot.mixed(wine.cor, lower = "ellipse", upper = "number", tl.pos = "lt", diag = "l", tl.col = "black")
```

```{r}
corrplot.mixed(wine.cor, lower = "ellipse", upper = "number", tl.pos = "lt", diag = "l", tl.col = "black")
```

#### 3.5.4 Combining corrgram with the significance test

```{r}
wine.sig = cor.mtest(wine.cor, conf.level= .95)
```

```{r}
corrplot(wine.cor, method = "number", type = "lower", diag = FALSE,
tl.col = "black", tl.srt = 45, p.mat = wine.sig$p, sig.level = .05)
```

#### 3.5.5 Reorder a corrgram

```{r}
corrplot.mixed(wine.cor, lower = "ellipse", upper = "number", tl.pos = "lt", diag = "l", order="AOE", tl.col = "black")
```

#### 3.5.6 Reordering a correlation matrix using hclust

```{r}
corrplot(wine.cor, method = "ellipse", tl.pos = "lt", tl.col = "black", order="hclust", hclust.method = "ward.D", addrect = 3)
```

# Lesson 5-C: Heatmap for Visualising and Analysing Multivariate Data

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## 2.0 Data

Loading the data to the R environment by using read_csv() function of readr package

```{r}
wh <- read_csv("chap05/data/whdata2018.csv")
```

Change the rows by country name instead of row number by using the code chunk below:

```{r}
wh$Country <- make.unique(as.character(wh$Country))
row.names(wh) <- wh$Country
```

The data was loaded into a data frame, but it has to be a data matrix to make a heatmap. The code chunk below will be used to transform wh data frame into a data matrix.

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh1)
```

## 3.0 Static Heatmap

### 3.1 heatmap() of R Stats

```{r}
wh_heatmap <- heatmap(wh_matrix, Rowv=NA, Colv=NA)
```

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

```{r}
wh_heatmap <- heatmap(wh_matrix, scale="column", cexRow = 0.6, cexCol = 0.8, margins = c(10, 4))
```

### 3.2 Creating Interactive Heatmap

```{r}
heatmaply(mtcars)
```

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

#### 3.2.1 Scaling method

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)], scale = "column")
```

#### 3.2.2 Normalising method

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

#### 3.2.3 Percentising method

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### 3.3 Clustering Algorithm

#### 3.3.1 Manual Approach

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), dist_method = "euclidean", hclust_method = "ward.D")
```

#### 3.3.2 Statistical Approach

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), dist_method = "euclidean", hclust_method = "average", k_row = 3)
```

#### 3.3.3 Seriation

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "OLO")
```

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "GW")
```

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "mean")
```

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "none")
```

#### 3.3.4 Working with color palettes

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), seriate = "none", colors = Blues)
```

#### 3.3.5 The Finishing touch

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]), Colv=NA, seriate = "none", colors = Blues, k_row = 5, margins = c(NA,200,60,NA), fontsize_row = 4, fontsize_col = 5, main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method", xlab = "World Happiness Indicators", ylab = "World Countries")
```


# Lesson 5-D: Treemap Visualisation with R

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(treemap, treemapify, tidyverse) 
```

## 2.0 Data

Loading the data to the R environment by using read_csv() function of readr package

```{r}
realis2018 <- read_csv("chap05/data/realis2018.csv")
```

## 3.0 Data Wrangling and Manipulation
### 3.1 Grouped summaries without the Pipe
```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`, `Planning Region`, `Planning Area`, `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped, `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), `Total Area` = sum(`Area (sqm)`, na.rm = TRUE), `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE), `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

### 3.2 Grouped summaries with the Pipe
```{r}
realis2018_summarised <- realis2018 %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

## 4.0 Designing Treemap with treemap Package
### 4.1 Designing a static treemap
```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

### 4.2 Using the basic arguments
```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 4.3 Working with vColor and type arguments
```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 4.4 Colours in treemap package
#### 4.4.1 The “value” type treemap
```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

#### 4.4.1 The “manual” type treemap
```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```


#### 4.4.2 Treemap Layout - working with alogorithm argument
```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

#### 4.4.3 Treemap Layout - using sortID
```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 4.5 Designing Treemap using treemapify Package
#### 4.5.1 Designing a basic treemap
```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") + 
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

#### 4.5.2 Defining hierarchy
```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") + 
  geom_treemap()
```

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap()
```

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

### 4.6 Designing Interactive Treemap using d3treeR
#### 4.6.1 Installing d3treeR package
```{r}
library(devtools)
install_github("timelyportfolio/d3treeR")
```

```{r}
library(d3treeR)
```

#### 4.6.2 Designing An Interactive Treemap
```{r}
tm <- treemap(realis2018_summarised,
        index=c("Planning Region", "Planning Area"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        title="Private Residential Property Sold, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

```{r}
d3tree(tm,rootname = "Singapore" )
```