---
title: "Hands-on Exercise 8"
author: "Teong Sze Ching"
date: "February 28, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Lesson 8A: Choropleth Mapping with R

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 2.0 Data

### 2.1 Importing Geospatial Data into R

The code chunk below uses the st_read() function of sf package to import MP14_SUBZONE_WEB_PL shapefile into R as a simple feature data frame called mpsz.

```{r}
mpsz <- st_read(dsn = "chap08/data", layer = "MP14_SUBZONE_WEB_PL")

mpsz
```

### 2.2 Importing Attribute Data into R

Next, import respopagsex2011to2020.csv file into RStudio and save the file into an R dataframe called popagsex.

```{r}
popdata <- read_csv("chap08/data/respopagesextod2011to2020.csv")
```

## 3.0 Data Preparation

Prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

YOUNG: age group 0 to 4 until age group 20 to 24, ECONOMY ACTIVE: age group 25-29 until age group 60-64, AGED: age group 65 and above, TOTAL: all age group, and DEPENDENCY: the ratio between young and aged against economy active group

### 3.1 Data wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

### 3.2 Joining the attribute data and geospatial data

Before performing the georelational join, one extra step is required to convert the values in PA and SZ fields to uppercase. This is because the values of PA and SZ fields are made up of upper- and lowercase. On the other, hand the SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, left_join() of dplyr is used to join the geographical data and attribute table using planning subzone name e.g. SUBZONE_N and SZ as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020, by = c("SUBZONE_N" = "SZ"))
```

Creating output:

```{r}
write_rds(mpsz_pop2020, "chap08/data/mpszpop2020.rds")
```

## 4.0 Choropleth Mapping Geospatial Data Using tmap

### 4.1 Plotting a choropleth map quickly by using qtm()

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

### 4.2 Creating a choropleth map by using tmap’s elements

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(fill = "DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.blues"),fill.legend = tm_legend(title = "Dependency ratio")) + tm_title("Distribution of Dependency Ratio by planning subzone") + tm_layout(frame = TRUE) + tm_borders(fill_alpha = 0.5) + tm_compass(type="8star", size = 2) + tm_grid(alpha =0.2) + tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", position = c("left", "bottom"))
```

#### 4.2.1 Drawing a base map

```{r}
tm_shape(mpsz_pop2020) + tm_polygons()
```

#### 4.2.2 Drawing a choropleth map using tm_polygons()

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY")
```

#### 4.2.3 Drawing a choropleth map using tm_fill() and tm_border()

```{r}
tm_shape(mpsz_pop2020)+ tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(fill = "DEPENDENCY") + tm_borders(lwd = 0.01, fill_alpha = 0.1)
```

### 4.3 Data classification methods of tmap

#### 4.3.1 Plotting choropleth maps with built-in classification methods

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "jenks", n = 5)) + tm_borders(fill_alpha = 0.5)
```

In the code chunk below, equal data classification method is used.

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "equal", n = 5)) + tm_borders(fill_alpha = 0.5)
```

#### 4.3.2 Plotting choropleth map with custome break

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) + tm_borders(fill_alpha = 0.5)
```

### 4.4 Colour Scheme

#### 4.4.1 Using ColourBrewer palette

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.greens")) + tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, add a “-” prefix.

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "-brewer.greens")) + tm_borders(fill_alpha = 0.5)
```

### 4.5 Map Layouts

#### 4.5.1 Map Legend

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons("DEPENDENCY", fill.scale = tm_scale_intervals(style = "jenks", n = 5, values = "brewer.greens"), fill.legend = tm_legend(title = "Dependency ratio")) + tm_borders(fill_alpha = 0.5) + tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

#### 4.5.2 Map style

```{r}
tm_shape(mpsz_pop2020)+ tm_fill("DEPENDENCY", style = "quantile", palette = "-Greens") + tm_borders(alpha = 0.5) + tmap_style("classic")
```

#### 4.5.3 Cartographic Furniture

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(fill = "DEPENDENCY", fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.blues"), fill.legend = tm_legend(title = "Dependency ratio")) + tm_title("Distribution of Dependency Ratio by planning subzone") + tm_layout(frame = TRUE) + tm_borders(fill_alpha = 0.5) + tm_compass(type="8star", size = 2) + tm_grid(alpha =0.2) + tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", position = c("left", "bottom"))
```

To reset the default style, refer to the code chunk below.

```{r}
tmap_style("white")
```

### 4.6 Drawing Small Multiple Choropleth Maps

#### 4.6.1 By assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+ tm_fill(c("YOUNG", "AGED"), style = "equal", palette = "Blues") + tm_layout(legend.position = c("right", "bottom")) + tm_borders(alpha = 0.5) + tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+ tm_polygons(c("DEPENDENCY","AGED"), style = c("equal", "quantile"), palette = list("Blues","Greens")) + tm_layout(legend.position = c("right", "bottom"))
```

#### 4.6.2 By defining a group-by variable in tm_facets()

```{r}
tm_shape(mpsz_pop2020) + tm_fill("DEPENDENCY", style = "quantile",
 palette = "Blues", thres.poly = 0) + tm_facets(by="REGION_N",  free.coords=TRUE) + tm_layout(legend.show = FALSE, title.position = c("center", "center"), title.size = 20) + tm_borders(alpha = 0.5)
```

#### 4.6.3 By creating multiple stand-alone maps with tmap_arrange()

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ tm_polygons("YOUNG", style = "quantile", palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ tm_polygons("AGED", style = "quantile", palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

#### 4.6.4 Mappping Spatial Object Meeting a Selection Criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
tm_fill("DEPENDENCY", style = "quantile", palette = "Blues", legend.hist = TRUE, legend.is.portrait = TRUE, legend.hist.z = 0.1) + tm_layout(legend.outside = TRUE, legend.height = 0.45, legend.width = 5.0, legend.position = c("right", "bottom"), frame = FALSE) + tm_borders(alpha = 0.5)
```

# Lesson 8B: Visualising Geospatial Point Data

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 2.0 Data

Import dataset..

```{r}
sgpools <- read_csv("chap08/data/SGPools_svy21.csv")
```

```{r}
list(sgpools) 
```

## 3.0 Creating a sf data frame from an aspatial data frame

The code chunk below converts sgpools data frame into a simple feature data frame by using st_as_sf() of sf packages

```{r}
sgpools_sf <- st_as_sf(sgpools, coords = c("XCOORD", "YCOORD"), crs= 3414)
```

```{r}
list(sgpools_sf)
```

## 4.0 Drawing Proportional Symbol Map

```{r}
tmap_mode("view")
```

### 4.1 It all started with an interactive point symbol map

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "red", size = 1, col = "black", lwd = 1)
```

To draw a proportional symbol map, we need to assign a numerical variable to the size visual attribute. The code chunks below show that the variable Gp1Gp2Winnings is assigned to size visual attribute.

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "red", size = "Gp1Gp2 Winnings", col = "black", lwd = 1)
```

### 4.2 Lets give it a different colour

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "OUTLET TYPE", size = "Gp1Gp2 Winnings", col = "black", lwd = 1)
```

### 4.2 twin map

```{r}
tm_shape(sgpools_sf) + tm_bubbles(fill = "OUTLET TYPE", size = "Gp1Gp2 Winnings", col = "black", lwd = 1) + tm_facets(by= "OUTLET TYPE", nrow = 1, sync = TRUE)
```

```{r}
tmap_mode("plot")
```

# Lesson 8C: Analytical Mapping

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 2.0 Data

Import dataset..

```{r}
NGA_wp <- read_rds("chap08/data/NGA_wp.rds")
```

## 3.0 Basic Choropleth Mapping

```{r}
p1 <- tm_shape(NGA_wp) + tm_polygons(fill = "wp_functional", fill.scale = tm_scale_intervals(style = "equal", n = 10, values = "brewer.blues"), fill.legend = tm_legend(position = c("right", "bottom"))) + tm_borders(lwd = 0.1, fill_alpha = 1) + tm_title("Distribution of functional water point by LGAs")
```

```{r}
p2 <- tm_shape(NGA_wp) + tm_polygons(fill = "total_wp", fill.scale = tm_scale_intervals(style = "equal", n = 10, values = "brewer.blues"), fill.legend = tm_legend(position = c("right", "bottom"))) + tm_borders(lwd = 0.1, fill_alpha = 1) +  tm_title("Distribution of total  water point by LGAs")
```

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

## 4.0 Choropleth Map for Rates

### 4.1 Deriving Proportion of Functional Water Points and Non-Functional Water Points

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

### 4.2 Plotting map of rate

```{r}
tm_shape(NGA_wp) + tm_polygons("pct_functional", fill.scale = tm_scale_intervals(style = "equal", n = 10, values = "brewer.blues"), fill.legend = tm_legend(position = c("right", "bottom"))) + tm_borders(lwd = 0.1, fill_alpha = 1) + tm_title("Rate map of functional water point by LGAs")
```

## 5.0 Extreme Value Maps

### 5.1 Percentile Map

#### 5.1.1 Data Preparation

Step 1: Exclude records with NA by using the code chunk below.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2: Creating customised classification and extracting values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

#### 5.1.2 Creating the get.var function

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### 5.1.3 A percentile mapping function

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){percent <- c(0,.01,.1,.5,.9,.99,1)
var <- get.var(vnam, df)
bperc <- quantile(var, percent)
tm_shape(df) + tm_polygons() + tm_shape(df) + tm_polygons(vnam, title=legtitle, breaks=bperc, palette="Blues", labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  + tm_borders() + tm_layout(main.title = mtitle, title.position = c("right","bottom"))
}
```

#### 5.1.4 Test drive the percentile mapping function

```{r}
percentmap("total_wp", NGA_wp)
```

### 5.2 Box map

```{r}
ggplot(data = NGA_wp, aes(x = "", y = wp_nonfunctional)) + geom_boxplot()
```

#### 5.2.1 Creating the boxbreaks function

```{r}
boxbreaks <- function(v,mult=1.5) {
qv <- unname(quantile(v))
iqr <- qv[4] - qv[2]
upfence <- qv[4] + mult * iqr
lofence <- qv[2] - mult * iqr
# initialize break points vector
bb <- vector(mode="numeric",length=7)
# logic for lower and upper fences
if (lofence < qv[1]) {  # no lower outliers
  bb[1] <- lofence
  bb[2] <- floor(qv[1])
} else {
  bb[2] <- lofence
  bb[1] <- qv[1]
}
if (upfence > qv[5]) { # no upper outliers
  bb[7] <- upfence
  bb[6] <- ceiling(qv[5])
} else {
  bb[6] <- upfence
  bb[7] <- qv[5]
}
bb[3:5] <- qv[2:4]
return(bb)
}
```

#### 5.2.2 Creating the get.var function

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### 5.2.3 Test drive the newly created function

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

#### 5.2.4 Boxmap function

```{r}
boxmap <- function(vnam, df, legtitle=NA, mtitle="Box Map", mult=1.5){
var <- get.var(vnam,df)
bb <- boxbreaks(var)
tm_shape(df) + tm_polygons() + tm_shape(df) + tm_fill(vnam,title=legtitle, breaks=bb, palette="Blues", labels = c("lower outlier", "< 25%", "25% - 50%", "50% - 75%", "> 75%", "upper outlier")) + tm_borders() + tm_layout(main.title = mtitle,  title.position = c("left", "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
