---
title: "Hands-on Exercise 9"
author: "Teong Sze Ching"
date: "March 7, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Lesson 9: Modelling, Visualising and Analysing Network Data with R

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(igraph, tidygraph, ggraph, visNetwork, lubridate, clock, tidyverse, graphlayouts, concaveman, ggforce)
```

## 2.0 Data

### 2.1 Importing Data

The data sets used in this hands-on exercise is from an oil exploration and extraction company. There are two data sets. One contains the nodes data and the other contains the edges (also know as link) data.

Edges data: GAStech-email_edges.csv which consists of two weeks of 9063 emails correspondances between 55 employees.

Nodes data: GAStech_email_nodes.csv which consist of the names, department and title of the 55 employees.

```{r}
GAStech_nodes <- read_csv("chap09/data/GAStech_email_node.csv")
GAStech_edges <- read_csv("chap09/data/GAStech_email_edge-v2.csv")
```

### 2.2 Reviewing the imported data

Examine the structure of the data frame using glimpse() of dplyr.

```{r}
glimpse(GAStech_edges)
```

### 2.3 Wrangling time

The output report of GAStech_edges above reveals that the SentDate is treated as “Character” data type instead of date data type. This is an error! Before we continue, it is important for us to change the data type of SentDate field back to “Date”” data type.

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate, label = TRUE, abbr = FALSE))
```

### 2.4 Wrangling attributes

A close examination of GAStech_edges data.frame reveals that it consists of individual e-mail flow records. This is not very useful for visualisation. In view of this, we will aggregate the individual by date, senders, receivers, main subject and day of the week.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

## 3.0 Creating network objects using tidygraph

### 3.1 Using tbl_graph() to build tidygraph data model

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes, edges = GAStech_edges_aggregated, directed = TRUE)
```

### 3.2 Changing the active object

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

## 4.0 Plotting Static Network Graphs with ggraph package

### 4.1 Plotting a basic network graph

```{r}
ggraph(GAStech_graph) + geom_edge_link() + geom_node_point()
```

### 4.2 Changing the default network graph theme

```{r}
g <- ggraph(GAStech_graph) + geom_edge_link(aes()) + geom_node_point(aes())

g + theme_graph()
```

### 4.3 Changing the coloring of the plot

```{r}
g <- ggraph(GAStech_graph) + geom_edge_link(aes(colour = 'grey50')) + geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10', text_colour = 'white')
```

### 4.4 Fruchterman and Reingold layout

```{r}
g <- ggraph(GAStech_graph, layout = "fr") + geom_edge_link(aes()) +
geom_node_point(aes())

g + theme_graph()
```

### 4.5 Modifying network nodes

```{r}
g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes()) + geom_node_point(aes(colour = Department,  size = 3))

g + theme_graph()
```

### 4.6 Modifying edges

```{r}
g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) +  scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 3)

g + theme_graph()
```

## 5.0 Creating facet graphs

### 5.1 Working with facet_edges()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2)

g + facet_edges(~Weekday)
```

The code chunk below uses theme() to change the position of the legend.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2) + theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

### 5.2 A framed facet graph

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, layout = "nicely") +  geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2)
  
g + facet_edges(~Weekday) + th_foreground(foreground = "grey80", border = TRUE) + theme(legend.position = 'bottom')
```

### 5.3 Working with facet_nodes()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, layout = "nicely") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department), size = 2)
  
g + facet_nodes(~Department)+ th_foreground(foreground = "grey80", border = TRUE) + theme(legend.position = 'bottom')
```

## 6.0 Network Metrics Analysis

### 6.1 Computing centrality indices

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") +  geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department, size=betweenness_centrality))
g + theme_graph()
```

### 6.2 Visualising network metrics

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(colour = Department, size = centrality_betweenness()))
g + theme_graph()
```

### 6.3 Visualising community

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") +  geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(
aes(colour = community))  

g + theme_graph()
```

```{r}
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(group_optimal(weights = Weight)),  betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") + geom_mark_hull(aes(x, y, group = community, fill = community), alpha = 0.2, expand = unit(0.3, "cm"), radius = unit(0.3, "cm")) + geom_edge_link(aes(width=Weight), alpha=0.2) + scale_edge_width(range = c(0.1, 5)) + geom_node_point(aes(fill = Department, size = betweenness_measure), color = "black", shape = 21)
  
g + theme_graph()
```

## 7.0 Building Interactive Network Graph with visNetwork

### 7.1 Data Preparation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

### 7.2 Plotting the first interactive network graph

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated)
```

### 7.3 Working with layout

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

### 7.4 Working with visual attributes - Nodes

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

When we rerun the code chunk below, visNetwork shades the nodes by assigning unique colour to each category in the group field

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### 7.5 Working with visual attributes - Edges

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", smooth = list(enabled = TRUE, type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123) 
```

### 7.6 Interactivity

```{r}
visNetwork(GAStech_nodes, GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
