---
title: "Take-Home Exercise 2"
author: "Teong Sze Ching"
date: "March 3, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Background

Singapore’s resident total fertility rate (TFR) has dropped to a historic low of 0.87 (extracted from CNA news article dated 26 Feb 2026), and policymakers have flagged low fertility and rapid ageing as a major national challenge.

However, fertility is influenced by multiple interacting factors (marriage patterns, divorce trends, and birth postponement across age groups), and these indicators are often published as separate series. 

This project builds an interactive visual analytics Shiny module to help users explore when key shifts occurred, which age bands drove fertility changes, and how strongly fertility co-moves with marriage/divorce trends (including possible lag effects).

## 1.0 Getting Started

Install and launch the following R packages

```{r}
pacman::p_load(tidyverse, readxl, patchwork, scales)
```

## 2.0 Data

Loading the data to the R environment. Dataset was extracted from CEIC database. 

```{r}

```

## 3.0 Data Preparation

To prepare the dataset for analysis and visualization, data cleaning and recoding steps were performed.


```{r}

```


------------------------------------------------------------------------

## 4.0 EDA

Goal of this is to understand temporal patterns before modelling.

Key Questions:
1. When did fertility begin declining?
2. Are marriage and fertility visually moving together?
3. Is fertility shifting toward older age groups?

:::: callout-note
## Visualization 1 - Total Fertility rate {.tabset}

::: panel-tabset
### Visual

```{r, echo=FALSE}

```

**Observation:**\
TBA

### Code

``` r

```
:::
::::

:::: callout-note
## Visualization 2 - Marriage rate {.tabset}

::: panel-tabset
### Visual

```{r, echo=FALSE}

```

**Observation:**\
xxx

### Code

``` r

```
:::
::::

:::: callout-note
## Visualization 3 - Divorce rate {.tabset}

::: panel-tabset
### Visual

```{r, echo=FALSE}

```

**Observation:**\
xxx

### Code

``` r

```
:::
::::

:::: callout-note
## Visualization 4 - Age-specific / Age group fertility rates  {.tabset}

::: panel-tabset
### Visual

```{r, echo=FALSE}

```

**Observation:**\
xxx

### Code

``` r

```
:::
::::

:::: callout-note
## Visualization 5 - Ageing population trend {.tabset}

::: panel-tabset
### Visual

```{r, echo=FALSE}


```

**Observation:**\
xxx

### Code

``` r

```
:::
::::


